---
title: "Introduction to R"
author: "Arun Frey"
institute: "Department of Sociology, University of Oxford"
date: "08.11.2021"
output:
  xaringan::moon_reader:
    #css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, warning=FALSE}

library(xaringanthemer)
library(knitr)
library(tidyverse)
library(png)

#source("hooks.R")
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, cache = TRUE)

#style_mono_accent(
#  base_color = "#1c5253",
#  header_font_google = google_font("Josefin Sans"),
#  text_font_google   = google_font("Montserrat", "300", "300i"),
#  code_font_google   = google_font("Fira Mono")
#)

theme_set(theme_bw())

starwars <- data.frame(starwars)

```

class: inverse, center, middle

# Recap of yesterday

---
# Vectors
```{r}
# vectors
x <- 1
x2 <- c(1, 2, 1, 1)
x3 <- c("hello", "hello", "course")

# vector functions
length(x2)

unique(x3)

sum(x2)
round(mean(x2), digits = 0)

paste(unique(x3), collapse = " ")

class(x3)
```

---
# Vector indexing 

```{r}
normal <- rnorm(1000, mean = 0, sd = 1)

normal[1:3]

normal_pos <- normal

normal_pos[normal_pos<=0] <- NA
```

---

```{r}
par(mfrow=c(1,2))
hist(normal)
hist(normal_pos)
```

---
# Dataframes
```{r}
# dataframes
class_df <- data.frame(
  "name" = c("Arun", "Mark", "Jenn"),
  "age" =  c(29, 28, 25), 
  "dep" = c("soc", "soc", "pol"), 
  "degree" = c("dphil", "dphil", "msc")
)

# read data
##read.csv, foreign::read.dta, read.delim
```

---

```{r}
# inspect data
head(class_df, 2)

tail(class_df, 2)

str(class_df)

summary(class_df)
```

---
class: inverse, middle, center

# Questions? 

---
# Accessing variables in dataframes

To access the names of variables (i.e. columns) in our data, use `names()`. 

```{r}
data("starwars")
names(starwars)
```

---

# Selecting variables using `$`

You can select variables using the `$` operator. 


```{r}
star_head <- data.frame(head(starwars, 3))

star_head$height

# is (almost) the same as
star_head[, "height"]

class(star_head$height)

class(star_head[, "height"])

class(star_head[["height"]])
```

---

# Selecting variables using `$`

Using these variables, you can then use that variable for descriptive statistics, using the functions you've learnt

```{r}
mean(starwars$height, na.rm = T)

# show first 5 unique species
unique(starwars$species)[1:5]

head(starwars[, c("name", "species")])
```

---

# Adding new variables

```{r}
survey <- data.frame("index" = c(1, 2, 3),
                     "name" = c("Mark", "Arun", "Jenn"))

survey$age <- c(28, 29, 25)

survey

survey$age_sq <- survey$age^2
```

---
# Indexing in dataframes

A dataframe consists of `n * m` observations. 

- `n`: number of rows
- `m`: number of columns

Just like with indexing vectors, you can also select individual observations in a dataframe, by specifying the row and the column of the observation `df[rows, columns]`. 

```{r}
df <- data.frame("var1" = c(1, 2, 3), "var2" = c(4, 5, 6))

# observation in first row, first column
df[1, 1]

# observation in third row, second column
df[3, "var2"]

# first observation, all columns
df[1, ]

```

---
class: inverse

# Quiz

What does each of the following lines return? 

```{r, eval = F}

df[1, 2]

df[, 2]

df[1,1]

df[c(1, 2), 1]


```


---
class: inverse

# Quiz

What does each of the following lines return? 

```{r, eval = T}
df[1, 2]

df[, 2]

df[1,1]

df[c(1, 2), 1]

```


---
# Slicing data with [, ]

#### Logical vectors

You can also index dataframes using a logical vector

```{r}
# create a new dataframe with only
# those cars with 6 cylinders
mtcars_6cyl <- mtcars[mtcars$cyl==6, ]

# create a new dataframe of only cars with 6 cylinders and mpg above 20  
mtcars_6cyl_20mpg <- mtcars[mtcars$mpg>20 & mtcars$cyl==6]
```

---
# Slicing with subset

The `subset()` function allows you to slice data just like with `[]`, but the code is easier to write. 


```{r}
# subset mtcars to only cars with 6 cylinders and 20 or more mpg
subset(x = mtcars, 
       subset = cyl == 6 & 
         mpg > 20)

```

---
# Plotting data

You can plot descriptive statistics using the `plot()` and `hist()` commands. 

`hist()` visualises the distribution of one variable.

```{r, eval = F}
hist(starwars$height)
```

---

```{r, eval = T}
hist(starwars$height)
```
---

class: inverse

# Quiz

```{r}
star_df <- head(starwars, 10)

# calculate mean height

# proportion of humans in data

# generate a new variable that displays the mass per height
#of each starwars character
# store this in a new variable called mass_height
  
```
---

# Plotting data

If you want to visualise how two variables relate to one another, you can generate a scatterplot using `plot()`. In `plot()` you first specify the x-coordinates, and then the y-coordinates of the plot. A series of other optional arguments allow you to connect dots via lines, modify axes and labels etc. 

Using the `mtcars` dataset, which comes with base R, I can examine the relationship between a car's fuel consumption (in miles/gallon) and its weight. 

```{r, eval = F}
plot(x = mtcars$mpg, y = mtcars$wt)
```
---

```{r, eval = T}
plot(x = mtcars$mpg, y = mtcars$wt)
```

---
# Labelling a plot

I can add other, optional labels to the plot. Check `?plot` to see more options. 

```{r, eval = F}
plot(mtcars$mpg, mtcars$wt,
     main = "Correlation plot between a car's fuel consumption and weight",
     xlab = 'Miles per gallon',  
     ylab = 'Weight (in tons)', 
     pch = 16,                 
     col = gray(.0, .1))

```

---

```{r, eval = T, echo = F}
plot(mtcars$mpg, mtcars$wt,
     main = "Correlation plot between a car's fuel consumption and weight",
     xlab = 'Miles per gallon',  
     ylab = 'Weight (in tons)', 
     pch = 16,                 
     col = gray(.0, .6))

```

---
# Functions

So far we have been creating objects, and modifying objects through pre-specified functions. 

`R` also lets you write your own functions, however. 

Functions are useful because they allow you to automate repetitive processes, such as generating summary statistics, tables, or standardizing variables, for example. 

To create a function, you need to use the `function()` command.

---
# Functions

In the example below, we replicate the `mean` function, which calculates the arithmetric average of a numeric vector. 

```{r}
mean_ft <- function(x) {
  x_bar <- sum(x) / length(x)
  return(x_bar)
}
```

All objects stored within a function are internal to that function; they won't show up in your environment. `R` will only return those objects that you specifically ask to be returned. 

```{r}
age <- c(18, 20, 22)
mean_ft(age)
```

---
# If statements

What happens when we try to compute a mean of a character vector, using our newly built function? 

```{r, error = TRUE}
mean_ft(c("hello"))
```
We can add a more understandable error using the `if` function, which executes a function only if the `if()` statement is `TRUE`. In the example below, the `mean_ft` function will stop and throw the `Input must be numeric` error whenever a non-numeric vector is used as an input. 

```{r, error=TRUE}
mean_ft <- function(x) {
  if(!is.numeric(x)) {
    stop("Input must be numeric")
  } else {
    x_bar <- sum(x) / length(x)
    return(x_bar)
  }
}
```
---
# If statements

```{r, error=TRUE}
mean_ft("hello")

mean_ft(c(1, 2, 4))
```



---

class: inverse, middle, center

# Complete the Exercise worksheet



